Throughout the writing of this dissertation, I have received a great deal of support and assistance.

I would like to express my deep gratitude to Fábio Coelho and Ricardo Vilaça, my research supervisors. Without your assistance and dedicated involvement in every step throughout the process, this dissertation would not have been possible, and you brought this work to a higher level.

I would like to extend my sincere thanks to every member of High-Assurance Software Laboratory (HASLab) \& INESC TEC for all their help. You provided me with the tools that I needed to choose the right direction and successfully complete my dissertation.

I would like to express my profound gratitude to my family for investing in my education and always providing me everything I needed to strive, without which none of this work or the previous five years would have been possible. You are always there for me.

Some special words of gratitude go to all my friends who have always been a major
source of support when things would get a bit discouraging. The last few years would have been unthinkable without all of the stories we have shared and all the help you have provided.

The last word goes for my loved one, Mariana, whose invaluable and unconditional support was what kept me moving. Every time I was ready to quit, you did not let me, and I am forever grateful. This dissertation stands as a testament to your unfailing support and continuous encouragement.

\bigbreak
\hfill \textit{Thank you}
\bigbreak

%Fabio: Incluir esta referência para o financiamento Não esquecer o agradecimento ao INESC TEC e ao HASLab

This work is financed by National Funds through the Portuguese funding agency, FCT - Fundação
para a Ciência e a Tecnologia, within project UIDB/50014/2020.